Keep your Laziness in Check

### 2018-02-19 paper/main.tex
[paper] work in progress for section 2

\usepackage{xspace}
\usepackage{preamble}
\usepackage{relsize}
%\usepackage{mathpartir}

@@

In this section, we will discuss how to specify the laziness behaviors
of Haskell functions through running examples of purely functional
maps from integers to integers. The map operations of interest
are \lk{insert} and \lk{lookup} with the following types:
\begin{inlinecode}
insert :: IntMap -> Int -> Int -> IntMap Int
lookup :: IntMap -> Int        -> Maybe  Int
\end{inlinecode}
%
The \lk{IntMap} can be implemented with a few different laziness
flavors. We can choose a binary search tree representation for
efficient lookups, or we can choose an association list representation
for easy inserts. The binary search tree representation necessarily
evaluates the keys of the map to order the keys, but it can be lazy in
the values. On the other hand, the association list representation can
be lazy in the keys as well. To summarise, we have maps that are
strict in both keys and values, maps that are strict in keys but lazy
in values, and maps that are fully lazy in both keys and values.

@@

An evaluation of a lazy data structure can partially explore the lazy
structure's shape. For example, \lk{lookup} on an association list
based map evaluates the association list until it finds the first
matching key. If such a key is halfway inside the association list,
then the values past this key will not be evaluated in this operation.
%
To describe how much of a data structure is evaluated, we will
use \textit{demand types}. A \textit{demand type} is almost isomorphic
to the corresponding data type, except we will wrap all the
constructor arguments in a \lk{Thunk} to represent the possible lack
of evaluation. For example, consider the \lk{Tree} type and
its \textit{demand type} \lk{TreeD}.

>>>>

\usepackage{xspace}
\usepackage{preamble}
\usepackage{relsize}
\usepackage{float}
\usepackage{cleveref}
%\usepackage{mathpartir}

@@

In this section, we will discuss how to specify the laziness behaviors
of Haskell functions through running examples of purely functional
maps. The map operations of interest are \lk{insert} and \lk{lookup}
with the following types:
\begin{inlinecode}
insert :: IntMap -> Int -> Int -> IntMap Int
lookup :: IntMap -> Int        -> Maybe  Int
\end{inlinecode}
%
The \lk{IntMap} can be implemented with a few different laziness
flavors. We can choose a binary search tree representation for
efficient lookups, or we can choose an association list representation
for easy inserts. The binary search tree representation necessarily
evaluates the keys of the map to order the keys, but it has flexible
laziness behavior on the values of the map. On the other hand, the
association list representation is even more flexible since it can be
lazy in the keys as well. To summarise, we have maps that are strict
in both keys and values, maps that are strict in keys but lazy or
strict in values, and maps that are fully lazy in both keys and
values.

@@

An evaluation of a lazy data structure can partially explore the lazy
structure's shape. For example, \lk{lookup} on an association list
based map evaluates the association list until it finds the first
matching key. If such a key is halfway inside the association list,
then the map entries past this key will not be evaluated.
%
To describe how much of a data structure is evaluated, we will
use \textit{demand types}. A \textit{demand type} is almost isomorphic
to the corresponding data type, except we will wrap all the
constructor arguments in a \lk{Thunk} to represent the possible lack
of evaluation. For example, let the binary tree representation of maps
be the \lk{Tree} type and let us consider its \textit{demand
type} \lk{TreeD}.
\begin{inlinecode}
data Tree =              data TreeD =
  | Leaf                   | LeafD
  | Node Tree Int Tree     | NodeD (Thunk TreeD) (Thunk IntD) (Thunk TreeD)
\end{inlinecode}
%
Values of \lk{TreeD} type describe a shape of values in the \lk{Tree}
type, and values of \lk{Thunk} describe any prefix shape of a given
tree since it has the following definition:
\begin{inlinecode}
data Thunk a = E !a | T
\end{inlinecode}
%
The \lk{E} constructor stands for \textit{evaluated} and the \lk{T}
constructor stand for \textit{thunk}. In \Cref{fig:tree-treeD}, the
value \lk{treeShape} matches an evaluation that only demands the root
of \lk{tree} and the value in its left branch.

\begin{figure}[H]
\begin{minipage}{0.4\textwidth}
\begin{inlinecode}
tree = Node
         Node Leaf 1 Leaf
         2
         Node Leaf 3 Leaf
\end{inlinecode}
\end{minipage}
\begin{minipage}{0.4\textwidth}
\begin{inlinecode}
treeShape = NodeD
              E (NodeD T (E 1) T)
              E 2
              T
\end{inlinecode}
\end{minipage}
\caption{Example of tree shape}
\label{fig:tree-treeD}
\end{figure}
